Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-09 arxiv/conclusions.tex
wip
add conclusion

\section{Conclusions and Open Problems}

In this work, we analyzed the bias and variance of multi-task learning versus single-task learning.
We provided tight concentration bounds for the bias and the variance.
Based on these bounds, we analyzed the impact of three properties, including task similarity, sample size, and covariate shift on the bias and variance, to derive conditions for transfer.
%Each of these results (based on each of the 3 properties) gives a different algorithmic/practical insight.
We validated our theoretical results.
Based on the theory, we proposed to train multi-task models by incrementally adding labeled data and showed encouraging results inspired by our theory.
We describe several open questions for future work.
First, our bound on the bias term (cf. Lemma \ref{lem_cov_derivative}) involves an error term that scales down with $\rho_1$.
Tightening this error bound might cover the unexplained observations in Figure \ref{fig_model_shift_phasetrans}.
Second, it would be interesting to extend our results to non-linear settings.
We remark that this likely requires addressing significant technical challenges  to deal with non-linearity.

\iffalse
\newpage
\section*{Broader Impacts}

In this work, we provide a theoretical framework to help understand when multi-task learning performs well.
We approach this question by studying the bias-variance tradeoff of multi-task learning.
We provide new technical tools to analyze the bias and variance of multi-task learning.
%We relate the bounds to three properties of task data.
%Overall, our theoretical study provides a framework to help understand multi-task learning performance.
%We further provide guidance for detecting and mitigating negative transfer on image and text classification tasks.

Our theoretical framework has the potential to impact many other neighboring areas in the ML community.
(i) Our concentration bounds can apply to different settings such as soft parameter sharing \cite{R17} , kernel methods \cite{EMP05}, and convex formulation of multi-task learning \cite{ZY14}.
(ii) Our analysis of the bias-variance tradeoff can extend to transfer learning and domain adaptation \cite{K18}.
(iii) Our insights on positive and negative transfer can be useful in multimodal learning, where the data sources are usually heterogeneous.
(iv) Our fine-gained study on sample sizes have the potential to provide new insight in meta learning, where limited labeled samples presents a significant challenge.
Finally, since multi-task learning connects to a wide range of areas \cite{V20} such as semi-supervised learning, representation learning, and reinforcement learning \cite{YKGLHF20}.%
We believe that the tools we have developed and the framework we have provided can inspire followup works in these areas.

Our proposed algorithmic consequences also have the potential to help both researchers and practitioners to better develop their use cases of multi-task learning.
For one example, many medical applications use multi-task learning to train large-scale image classification models by combining multiple datasets \cite{chexnet17,EA20}.
Unlike the applications of multi-task learning in text classification where we can collect large amounts of labeled data \cite{GLUE}, in medical applications it is typically difficult and expensive to acquire labeled data.
For such settings, our proposed training scheduler might improve the model training efficiency by using less labeled data.
For another example, practitioners in industry often need to improve prediction performance by training many related tasks.
But the results are not always positive and practitioners have a hard time figuring out why.
Our proposed metric can provide guidance for selecting which tasks should be trained together in a multi-task model.
\fi

%Our algorithmic consequences of our theory have the potential to impact downstream applications of multi-task learning.
%For example, many medical applications use multi-task learning to train large-scale image classification models by combining multiple datasets \cite{chexnet17,EA20}.
%Unlike the applications of multi-task learning in text classification where large amounts of labeled data are collected \cite{GLUE}, in medical applications it is typically difficult to acquire large amounts of labeled data.
%For such settings, training multi-task models can be very challenging.
%Our insight on using single-task learning results to help understand multi-task learning can be valuable for helping practitioners understand their results.

>>>>

\section{Conclusions and Discussions}\label{sec_conclude}

This work proposed a formal study of a widely used hard paramaeter sharing approach in high-dimensional linear regression, where the sample size and feature dimension increase at a fixed ratio.
We provided tight generalization bounds that scale with dataset properties such as their sample sizes and covariate shifts.
Based on these bounds, we analyzed the impact of varying sample sizes and covariate shifts on the prediction loss of hard parameter sharing.
Our work rigorously explains several empirical phenomena such as sample efficiency and negative transfer related to these dataset properties.
We validated our results and conducted further studies on a real world classification task.

We describe several open questions for future work.
First, our bound in Corollary \eqref{cor_MTL_loss} involves an error term that scales down with $n_1$.
Tightening this error bound requires showing the limit of $\normFro{({Z^{(1)}}^{\top} Z^{(1)} + {Z^{(2)}}^{\top} Z^{(2)})^{-1} {Z^{(1)}}^{\top} Z^{(1)}}^2$ for two random matrices $Z^{(1)}, Z^{(2)}$ with isotropic covariances.
This might require new techniques beyond \HZ{to add}.
Second, it would be interesting to extend our results to classification problems.
Several recent work have made remarkable progress in the (one-sample) high-dimensional logistic regression setting \cite{sur2019modern}.
An interesting question is to study hard parameter sharing in (multiple-sample) high-dimensional logistitic regression.

### 2020-10-09 arxiv/experiments.tex
add conclusion

We validate our  algorithmic insights and then our theory.
In Section \ref{sec_exp_res}, we first show that single-task learning results can help predict positive or negative transfer.
%the single-task based metric on sentiment analysis and ChestX-ray14 datasets.
Second, our proposed incremental training schedule improves the training efficiency of standard multi-task training on sentiment analysis tasks.
%Third, we measure the data efficiency ratio of multi-task learning on six sentiment analysis tasks.
In Section \ref{sec_exp_ab}, we validate our theoretical results.
We further show that when the sample ratio is large, performing the alignment procedure of \cite{WZR20} provides more improvement for MTL.

@@

\subsection{Further Studies on a Text Classification Task}\label{sec_text}

Our results and their implications are all in the high-dimensional linear regression setting.
How well do they extend to other settings?
In this section, we conduct further studies on a text classification task, for predicting whether  a sentence has a positive or negative sentiment.
Our datasets include a review sentiment dataset (MR) \cite{pang2005seeing}, a sentence subjectivity dataset (SUBJ) \cite{pang2004sentimental}, a customer reviews dataset (CR) \cite{hu2004mining}, a question type dataset (TREC) \cite{li2002learning}, an opinion polarity dataset (MPQA) \cite{wiebe2005annotating}, and the Stanford sentiment treebank (SST) dataset \cite{socher2013recursive}.
%The question is to predict positive or negative sentiment expressed in the text.
Our model consists of a word embedding layer with GloVe embeddings \cite{pennington2014glove} followed by a long-short term memory (LSTM) or a multi-layer perception (MLP) layer \citet{lei2018simple}.\footnote{For multi-layer perception (MLP), we apply an average pooling layer over the word embeddings. For LSTM and CNN, we add a shared feature representation layer on top of the word embeddings.}

@@

\paragraph{Sample ratio.}
In Figure \ref{fig_ab_data}, we observe that for multiple example task pairs, increasing task one's sample size improves task two's prediction accuracy initially, but hurts eventually.
On the $y$-axis, we plot task two's test accuracy using HPS, subtracted by its STL test accuracy.
%validate that as we increase the sample ratio while keeping task two's sample size fixed, task two's prediction accuracy does not always increase.
We fix task one's sample size at $1000$ and increase task two's sample size from $100$ to $3000$.
These examples and the one in Figure \ref{fig_size} suggest a natural progressive training schedule, where we add samples progressively until performance drops.
%In particular, Figure \ref{fig_size} (and our analysis) shows that $L(\hat{\beta}_t^{\MTL})$ behaves as a quadratic function over $\rho_1$.
%More generally, depending on how large $\Psi(\beta_1, \beta_2)$ is, $L(\hat{\beta}_t^{\MTL})$ may also be monotonically increasing or decreasing.
%Based on this observation, we propose a progressive training schedule to improve the compuational efficiency of hard parameter sharing.
Concretely,
\begin{itemize}
	\item we divide the source task data into $S$ batches.
	For $S$ rounds, we incrementally add the source task data by adding one batch at a time.
	\item After training $T$ epochs, if the validation accuracy becomes worse than the previous round's result, we terminate.
%	Algorithm \ref{alg_inc_train} in Appendix \ref{app_experiments} describes the procedure in detail.
\end{itemize}
For example, if we apply this procedure to the settings of Figure \ref{fig_ab_data} and \ref{fig_size}, it will terminate until reaching the optimal sample ratio.
The advantage of this procedure is that it reduces the computational cost compared to standard round-robin training schedules.

>>>>

%We validate our  algorithmic insights and then our theory.
%In Section \ref{sec_exp_res}, we first show that single-task learning results can help predict positive or negative transfer.
%the single-task based metric on sentiment analysis and ChestX-ray14 datasets.
%Second, our proposed incremental training schedule improves the training efficiency of standard multi-task training on sentiment analysis tasks.
%Third, we measure the data efficiency ratio of multi-task learning on six sentiment analysis tasks.
%In Section \ref{sec_exp_ab}, we validate our theoretical results.
%We further show that when the sample ratio is large, performing the alignment procedure of \cite{WZR20} provides more improvement for MTL.

@@

\subsection{Further Studies on Text Classification Tasks}\label{sec_text}

Our results and their implications are all in the high-dimensional linear regression setting.
How well do they extend to other scenarios?
In this section, we conduct further studies on six text classification datasets, for predicting whether a sentence has a positive or a negative sentiment.
Our datasets include a review sentiment dataset (MR) \cite{pang2005seeing}, a sentence subjectivity dataset (SUBJ) \cite{pang2004sentimental}, a customer reviews dataset (CR) \cite{hu2004mining}, a question type dataset (TREC) \cite{li2002learning}, an opinion polarity dataset (MPQA) \cite{wiebe2005annotating}, and the Stanford sentiment treebank (SST) dataset \cite{socher2013recursive}.
%The question is to predict positive or negative sentiment expressed in the text.
Our model consists of a word embedding layer with GloVe embeddings \cite{pennington2014glove} followed by a long-short term memory (LSTM) or a multi-layer perception (MLP) layer \cite{lei2018simple}.\footnote{For MLP, we apply an average pooling layer over word embeddings. For LSTM, we add a shared feature representation layer on top of word embeddings.}

@@

\paragraph{Sample ratio.}
In Figure \ref{fig_ab_data}, we observe that for multiple example task pairs, increasing task one's sample size improves task two's prediction accuracy initially, but hurts eventually.
On the $y$-axis, we plot task two's test accuracy using HPS, subtracted by its STL test accuracy.
%validate that as we increase the sample ratio while keeping task two's sample size fixed, task two's prediction accuracy does not always increase.
We fix task one's sample size at $1000$ and increase task two's sample size from $100$ to $3000$.
These examples and the one in Figure \ref{fig_size} suggest a natural progressive training schedule, where we add samples progressively until performance drops:
%In particular, Figure \ref{fig_size} (and our analysis) shows that $L(\hat{\beta}_t^{\MTL})$ behaves as a quadratic function over $\rho_1$.
%More generally, depending on how large $\Psi(\beta_1, \beta_2)$ is, $L(\hat{\beta}_t^{\MTL})$ may also be monotonically increasing or decreasing.
%Based on this observation, we propose a progressive training schedule to improve the compuational efficiency of hard parameter sharing.
\begin{itemize}
	\item We divide the source task data into $S$ batches.
	For $S$ rounds, we incrementally add the source task data by adding one batch at a time.
	\item After training $T$ epochs, if the validation accuracy becomes worse than the previous round's result, we terminate.
%	Algorithm \ref{alg_inc_train} in Appendix \ref{app_experiments} describes the procedure in detail.
\end{itemize}
For example, if we apply this procedure to the settings of Figure \ref{fig_ab_data} and \ref{fig_size}, it will terminate until reaching the optimal sample ratio.
The advantage of this procedure is that it reduces the computational cost compared to standard round-robin training schedules.

### 2020-10-09 arxiv/intro.tex
section 3 4 polish writing
add conclusion

\begin{abstract}
	Hard parameter sharing is a widely used approach to learn from multiple tasks in many applications, such as text classification. The idea is that all tasks share the same feature layer, while each task uses a specific output layer for prediction. Intuitively, the performance of hard parameter sharing depends on dataset properties such as their sample sizes. Yet, rigorously formulating such intuition can be technically challenging. This paper analyzes the generalization performance of hard parameter sharing in high-dimensional linear regression, where the sample size and feature dimension become increasingly large in a fixed ratio. We present tight generalization bounds for two canonical cases: (i) multiple tasks with the same feature covariates; (ii) two tasks with arbitrarily different sample sizes and covariance matrices. We also demonstrate that these estimates for the empirical loss of hard parameter sharing are incredibly accurate in modest dimensions. Finally, we explain several intriguing empirical phenomena. For example, increasing one task's sample size helps another task initially by reducing variance, but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in a text classification task.
%	Multi-task learning is a powerful approach in many applications such as image and text classification. Yet, there is little rigorous understanding of when multi-task learning outperforms single-task learning. In this work, we provide a rigorous study to answer the question in the high-dimensional linear regression setting. We show that a bias-variance tradeoff of multi-task learning determines the effect of information transfer, and develop new concentration bounds to analyze the tradeoff. Our key observation is that three properties of task data, namely \textit{task similarity}, \textit{sample ratio}, and \textit{covariate shift} can affect transfer in the high-dimensional linear regression setting. We relate each property to the bias and variance of multi-task learning and explain three negative effects with decreased task similarity,	increased sample ratio, and covariate shift under increased sample ratio. We validate the three effects on text classification tasks. Inspired by our theory, we show two practical connections of interest.
%	First, single-task results can help to understand when multi-task learning gives gains. Second, incrementally adding training data can mitigate negative transfer and improve multi-task training efficiency.
\end{abstract}

@@

%Let $n_i = \rho_i \cdot p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
%The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We consider a multi-task estimator obtained using a shared linear layer for all tasks and a separate output layer for each task \cite{WZR20}.
This two-layer model is inspired by a commonly used idea of hard parameter sharing in multi-task learning \cite{R17,MTDNN19}.
We consider the bias and variance of the multi-task estimator for predicting a target task and compare its performance to single-task learning.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
%Let $\hat{\beta}_t^{\MTL}$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_prelim}.
%We revisit the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$.
\fi

>>>>

\begin{abstract}
	Hard parameter sharing is a widely used approach to learn from multiple tasks in many applications, such as text classification. The idea is that all tasks share the same feature layer, while each task uses a specific output layer for prediction. Intuitively, the performance of hard parameter sharing depends on dataset properties such as their sample sizes. Yet, rigorously formulating such intuition can be technically challenging. This paper analyzes the generalization performance of hard parameter sharing in high-dimensional linear regression, where the sample size and feature dimension become increasingly large in a fixed ratio. We present tight generalization bounds for two canonical cases: (i) multiple tasks with the same feature covariates; (ii) two tasks with arbitrarily different sample sizes and covariance matrices. We also demonstrate that these estimates for the empirical loss of hard parameter sharing are incredibly accurate in modest dimensions. Finally, we explain several intriguing empirical phenomena. For example, increasing one task's sample size helps another task initially by reducing variance, but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in text classification tasks.
%	Multi-task learning is a powerful approach in many applications such as image and text classification. Yet, there is little rigorous understanding of when multi-task learning outperforms single-task learning. In this work, we provide a rigorous study to answer the question in the high-dimensional linear regression setting. We show that a bias-variance tradeoff of multi-task learning determines the effect of information transfer, and develop new concentration bounds to analyze the tradeoff. Our key observation is that three properties of task data, namely \textit{task similarity}, \textit{sample ratio}, and \textit{covariate shift} can affect transfer in the high-dimensional linear regression setting. We relate each property to the bias and variance of multi-task learning and explain three negative effects with decreased task similarity,	increased sample ratio, and covariate shift under increased sample ratio. We validate the three effects on text classification tasks. Inspired by our theory, we show two practical connections of interest.
%	First, single-task results can help to understand when multi-task learning gives gains. Second, incrementally adding training data can mitigate negative transfer and improve multi-task training efficiency.
\end{abstract}

@@

%Let $n_i = \rho_i \cdot p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
%The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We consider a multi-task estimator obtained using a shared linear layer for all tasks and a separate output layer for each task \cite{WZR20}.
This two-layer model is inspired by a commonly used idea of hard parameter sharing in multi-task learning \cite{R17,MTDNN19}.
We consider the bias and variance of the multi-task estimator for predicting a target task and compare its performance to single-task learning.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
%Let $\hat{\beta}_t^{\MTL}$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_prelim}.
%We revisit the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$.

\HZ{go to intro}
For a multivariate Gaussian random matrix, this result follows from the classical result for the mean of inverse Wishart distribution \cite{anderson1958introduction}.
For a non-Gaussian random matrix, this result can be obtained using the well-known Stieltjes transform method (cf. Lemma 3.11 of \citet{bai2009spectral}).
\fi
